Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-15 writeup/conStruct.tex
tidbits

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

@@

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

@@

\include{macros}

@@

@@

From a practical standpoint, even while acknowledging continuous
processes, we often need to identify somewhat separable populations 
from which individuals are sampled \citep{wright1949genetical}.
Delineating populations is useful for systematics and for
informing conservation priorities \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.
Furthermore, we often need to identify subsets of indivduals resulting from reasonably coherent
evolutionary histories for downstream analyses to learn about population history and adaptation.
Conversely,
the substantial information available from continuous, geographic differentiation
(e.g., adaptation along a climatic gradient)
can be confounded by discrete historical processes (e.g., admixture),
requiring methods that can disentangle the two.

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \citep{CavalliSforza1975, treemix},
dimensionality-reduction approaches \citep{meirmans2009genodive}, 
such as principal components analysis 
\citep{menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
\citep[e.g.][]{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}, among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\citep{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster memberships \citep{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

\emph{Isolation by distance} refers to a pattern of increasing genetic differentiation
with geographic separation,
which occurs when geographically restricted dispersal allows
genetic drift to build up differentiation between distant locations
\citep{Wright1943}. 
Theoretical work,
mostly derived from ``stepping-stone'' models 
\citep{kimura1964stepping,sawyer1976stepping,shiga1984stepping},
gives us some analytical predictions for isolation by distance
\citep{malecot1969mathematics,Slatkin1985,epperson2003geographical}, 
but substantial work remains to be done \citep{barton2002neutral,barton2013modelling}.
% IBD has been modeled in both spatially continuous frameworks 
% in which an individual's parents are found within some radius of that individual's location,
% and discrete, ``stepping-stone'' frameworks \citep{kimura1964stepping},
% in which migration occurs only between neighboring demes.
Given the generality of the circumstances that generate a pattern of isolation by distance, 
it is unsurprising that IBD is very widespread in nature \citep{meirmans2012,Sexton_etal_2014}.

The ubiquity of isolation by distance presents a challenge for models of discrete population structure,
as it is frequently difficult to determine whether observed patterns of genetic variation are 
continuously distributed across a landscape, or instead are partitioned in discrete clusters.
This problem can be compounded if sampling is done unevenly or discretely across a population or species' range,
and has given rise to a debate in the population genetic literature
about how best to describe sets of individuals using continuous clines and discrete clusters 
\citep[e.g.][]{SerrePaabo2004,rosenberg2005clines}.

Existing model-based clustering approaches can only describe continuous patterns of variation using
discrete clusters, and so tend to erroneously describe continuously distributed variation with multiple clusters that 
show spatially autocorrelated cluster membership \citep{Frantz2009,meirmans2012}.
In analyses of empirical datasets, which often are characterized by IBD,
model-based clustering approaches will therefore tend to overestimate
the number of discrete clusters present.

@@

This definition of covariance differs from the usual ``genetic covariance'' \citep{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eq.\ \eqref{allelic_covariance} further in \secref{allelic_cov}.
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $\pi_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 \pi_{i,j})/4$.
Therefore, 
this allelic covariance is more affected in shape by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
if they are likely to contain a large percentage of errors.

@@

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each layer, 
we define the covariance within layer $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_layer_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th layer.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
and the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the layer.
Our choice of a powered-exponential decay, 
as parameterized by the $\alpha$s, 
is a flexible and standard choice in spatial statistics \citep{Diggle1998}, 
and is not chosen to match a particular population genetics model. 
% (Analytical theory \citep{barton2002neutral} predicts the decay should be proportional to $\exp{-D}/\sqrt{D}$.)
The $\phi^{(k)}$ is a parameter that describes the background covariance within the layer. 
If two samples draw 100\% of their ancestry from layer $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the layer.
The ``shared drift'' parameter $\phi^{(k)}$ is analogous to 
the branch length connecting the $k$th population to the population ancestral to all modeled
layers \citep[see for example ][]{patterson_ancient_2012,peter_fstats},
although they cannot be directly compared because 
we are modeling the allelic, rather than genetic, covariance. 
In \ref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

@@

\paragraph{Likelihood and inference}
If the allele frequencies at each locus
were independent between loci and multivariate normally distributed across populations, 
their allelic covariance $\widehat{\Omega}$ 
would be Wishart distributed with degrees of freedom equal to $L$, 
the number of loci genotyped.
We use this as a convenient approximation to the true distribution described above,
and so define the likelihood of the allelic covariance to be
\begin{equation}
    P(\widehat{\Omega} \; | \; \Omega) 
    = 
    \mathcal{W} \left( L\widehat{\Omega} \; | \; \Omega,L\right) ,
\end{equation}
where $\mathcal{W}$ is the Wishart likelihood function.
Statistical nonindependence between loci (linkage disequilibrium)
will decrease the effective
number of degrees of freedom.  
One possible solution, which we have not yet found necessary to implement,
would be to estimate an \emph{effective} number of loci 
by introducing a parameter to modify the given degrees of freedom 
and thereby informally model linkage between loci \citep[e.g.][]{EEMS}.

We estimate the values of the parameters of the model using a Bayesian approach.
Acknowledging the dependence of the parametric covariance matrix $\Omega$ on its constituent parameters
$w,\alpha,\phi,\eta,\gamma$ and on the (observed) geographic distances $D$ 
with the notation $\Omega(w,\alpha,\phi,\eta,\gamma,D)$,
we denote the posterior probability density of the parameters as:
\begin{equation}
P\left( w,\alpha,\phi,\eta,\gamma \;|\; \widehat{\Omega} \right) 
    \propto
    P\left(\widehat{\Omega} \; | \; \Omega(w,\alpha,\phi,\eta,\gamma,D) \right)
    P(w)P(\alpha)P(\phi)P(\eta)P(\gamma) ,
\end{equation}
where $P(w)$, $P(\alpha)$, $P(\phi)$, $P(\eta)$, and $P(\gamma)$, are prior distributions.
All parameters are given (half-)Gaussian priors 
except for $\alpha_2$, which is uniform on $(0,2)$, 
and $w$, for which we use an independent Dirichlet of dimension $K$ for each individual
(see Table \ref{tab:param_prior_tab} for specifics).
Parameters are independent between layers.
We use Hamiltonian Monte Carlo as implemented in STAN \citep{stan, NUTS, stan_lib, rstan} 
to estimate the posterior distribution on the parameters.
We also provide an R package \citep{R} called \texttt{conStruct} 
(for ``continuous structure") that functions as a wrapper around this inference machinery.

@@

A nice feature of our approach is that the model described in Eq.\ \eqref{cross_layer_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is very similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the \gc{main} difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \citep{Engelhardt2012}.  
(We make this approximation for the substantial advantages in
computational speed.)
\gc{The second small difference is that in the original STRUCTURE model,
  allele frequencies at each locus are independently drawn for each
  cluster \citep{STRUCTURE}, while in conStruct's non-spatial model,
  it is more natural to envision each layer's allele frequency
  as being drifted away from a single, global allele frequency. 
  This makes our model more
  closely related to the ``$F$-model'' prior for allele frequencies of \citet{falush2003}.
As these differences are relatively small,} we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework. 
We also validate our claim that the results of our nonspatial method 
are a close match to those of STRUCTURE-based approaches.

@@

Several approaches have been used as model choice criteria 
for the number of discrete clusters in population genetic data, including: 
information theoretic criteria \citep[e.g.][]{ADMIXTURE}, such as AIC or BIC;
comparisons of the likelihood of the data across different values of $K$, 
with various criteria on how to choose a single value (e.g., \citep{Evanno2005});
or comparisons of the marginal likelihood, 
generated either via various approximations \citep[e.g.][]{STRUCTURE}
or via a more rigorous and computationally intensive approach such as thermodynamic integration \citep{verity_nichols2016}
or inference using a Dirichlet process prior \citep{huelsenbeck2007inference}.
See \cite{verity_nichols2016} for a discussion of these approaches and comparison
between several methods.

We use cross-validation \citep[similar in spirit to][]{ADMIXTURE_xval} to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

@@

Another way to describe the practical significance of additional layers
is to calculate each layer's relative contribution to total covariance, 
and to choose a value of $K$ where all layers have a contribution above some cutoff (e.g. 0.1\%).
The Dirichlet prior on admixture proportions
is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
encouraging the model to ``not use" unnecessary layers if they are present, 
so that they will have a low contribution to overall covariance.

To calculate layer contributions, we use the following alternative description of our covariance model:
the genomes of any pair of individuals \emph{agree} with some background probability,
but this probability of agreement is increased on any segment of genome 
that both have inherited through the same layer 
(the amount it increases depends on how far apart they are.)
We use this characterization to quantify the relative contributions of each layer,
by computing the average contribution to increased probability of agreement
as described in \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

To test the method, we first generated data using the coalescent simulator \texttt{ms} \citep{Hudson2002}.
In each simulation,
we split a single ancestral population into $K$ subpopulations $\tau_{\text{s}}$ units of coalescent time in the past,
and at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonized a
separate $6 \times 6$ square lattice of demes.
Migration on each lattice was to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past, 
we collapsed those $K$ discrete layers into a single grid of demes,
choosing various amounts of admixture from these different layers
(see Fig \ref{sim_setup}). 
We collapsed the layers together using random admixture with spatially autocorrelated proportions 
(see \secref{sim_details} for details), 
so that both the frequencies in layers and the layer memberships are
spatially autocorrelated. 
We simulated datasets using $K=1$, 2, and 3 layers;
in each simulation we sampled 10,000 unlinked loci from each of 20 haploid individuals from every deme.
We then ran both spatial and nonspatial \texttt{conStruct} analyses on each simulated dataset with $K$ between 1 and 7,
and compared predictive performance of the models using cross-validation.
% To confirm that results from our nonspatial model were qualitatively similar to those of 
% other nonspatial model-based clustering algorithms, 
For comparison,
we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, using $K=2-4$.
% inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}--\ref{fastStr_simK3}).

% In Fig \ref{simK1_pies}, we show the results of the dataset simulated with $K=1$, 
% contrasting the results of the nonspatial model runs with 2--4 layers 
% with those of spatial model with 2--4 layers.
% 
% In Fig \ref{sim_xvals} we show the results of the cross-validation analysis for each of the three simulated datasets.
% The results of all analyses run on datasets simulated using $K=1$ through 3 are shown in 
% Figs \ref{simK1_nsp_pies}-\ref{simK3_adprop_fit}.
% 
% To confirm that results from our nonspatial model were qualitatively similar to those of 
% other nonspatial model-based clustering algorithms, 
% we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, 
% inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}--\ref{fastStr_simK3}).

@@

Trees in the genus \textit{Populus} (poplars, aspens, and cottonwoods) 
are distributed throughout the Northern Hemisphere;
species in the genus regularly co-occur and, 
where they do, they frequently hybridize \citep{eckenwalder1984, Cronk2005}.

\textit{Populus trichocarpa}, the black cottonwood,
and \textit{Populus balsamifera}, the balsam poplar,
have a broad zone of overlap in the Pacific Northwest,
where they are hypothesized to hybridize \citep{geraldes_etal_2014, suarezgonzalez_etal_2016}.
Both species are sampled over a large geographic region, 
and show spatial patterns of genetic and phenotypic variation \citep{slavov_etal_2012, mckown_etal_2013},
making the system well-suited for application of our method.
% to test an analysis designed to describe
% how genetic structure is partitioned both discretely across populations (or species)
% and continuously with geographic distance.
We organize the results of our analyses around the following questions: 
\begin{enumerate}

@@

Across all values of $K$ for which we ran \texttt{conStruct},
we see strong support for the spatial model over the nonspatial model (Fig \ref{bear_xvals}).
% indicating the isolation by distance seems to be a prominent feature of the data.
This pattern may resolve a discrepancy between our results and 
previous analyses that split Alaskan and British Columbian bears out 
into their own cluster with an inferred Beringian glacial refugium
\citep{Byun1997,Stone2000,Puckett2015}.
Our model, which explicitly incorporates a spatial decay of relatedness, 
allows somewhat genetically differentiated individuals 
that are sampled far from one another to belong to the same layer,
instead of splitting these individuals out into successive clusters 
(e.g., Fig \ref{bear_sp_pies}d vs \ref{bear_nsp_pies}d).

>>>>

% Use the PLoS provided BiBTeX style
\bibliographystyle{latex/plos2015}

@@

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{latex/PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

@@

@@

\newcommand{\R}{\mathbb{R}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\supp}{\mathop{\mbox{supp}}}
\newcommand{\sgn}{\mathop{\mbox{sgn}}}
\newcommand{\bone}{\mathbf{1}}

@@

From a practical standpoint, even while acknowledging continuous
processes, we often need to identify somewhat separable populations 
from which individuals are sampled \cite{wright1949genetical}.
Delineating populations is useful for systematics and for
informing conservation priorities \cite{Moritz1994,Waples_1998,Moritz_etal_2002}.
Furthermore, we often need to identify subsets of indivduals resulting from reasonably coherent
evolutionary histories for downstream analyses to learn about population history and adaptation.
Conversely,
the substantial information available from continuous, geographic differentiation
(e.g., adaptation along a climatic gradient)
can be confounded by discrete historical processes (e.g., admixture),
requiring methods that can disentangle the two.

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \cite{CavalliSforza1975, treemix},
dimensionality-reduction approaches \cite{meirmans2009genodive}, 
such as principal components analysis 
\cite{menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
(e.g., \cite{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}), among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\cite{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster memberships \cite{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

\emph{Isolation by distance} refers to a pattern of increasing genetic differentiation
with geographic separation,
which occurs when geographically restricted dispersal allows
genetic drift to build up differentiation between distant locations
\cite{Wright1943}. 
Theoretical work,
mostly derived from ``stepping-stone'' models 
\cite{kimura1964stepping,sawyer1976stepping,shiga1984stepping},
gives us some analytical predictions for isolation by distance
\cite{malecot1969mathematics,Slatkin1985,epperson2003geographical}, 
but substantial work remains to be done \cite{barton2002neutral,barton2013modelling}.
% IBD has been modeled in both spatially continuous frameworks 
% in which an individual's parents are found within some radius of that individual's location,
% and discrete, ``stepping-stone'' frameworks \cite{kimura1964stepping},
% in which migration occurs only between neighboring demes.
Given the generality of the circumstances that generate a pattern of isolation by distance, 
it is unsurprising that IBD is very widespread in nature \cite{meirmans2012,Sexton_etal_2014}.

The ubiquity of isolation by distance presents a challenge for models of discrete population structure,
as it is frequently difficult to determine whether observed patterns of genetic variation are 
continuously distributed across a landscape, or instead are partitioned in discrete clusters.
This problem can be compounded if sampling is done unevenly or discretely across a population or species' range,
and has given rise to a debate in the population genetic literature
about how best to describe sets of individuals using continuous clines and discrete clusters 
(e.g., \cite{SerrePaabo2004,rosenberg2005clines}).

Existing model-based clustering approaches can only describe continuous patterns of variation using
discrete clusters, and so tend to erroneously describe continuously distributed variation with multiple clusters that 
show spatially autocorrelated cluster membership \cite{Frantz2009,meirmans2012}.
In analyses of empirical datasets, which often are characterized by IBD,
model-based clustering approaches will therefore tend to overestimate
the number of discrete clusters present.

@@

This definition of covariance differs from the usual ``genetic covariance'' \cite{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eq.\ \eqref{allelic_covariance} further in \secref{allelic_cov}.
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $\pi_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 \pi_{i,j})/4$.
Therefore, 
this allelic covariance is more affected in shape by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
if they are likely to contain a large percentage of errors.

@@

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each layer, 
we define the covariance within layer $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_layer_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th layer.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
and the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the layer.
Our choice of a powered-exponential decay, 
as parameterized by the $\alpha$s, 
is a flexible and standard choice in spatial statistics \cite{Diggle1998}, 
and is not chosen to match a particular population genetics model. 
% (Analytical theory \cite{barton2002neutral} predicts the decay should be proportional to $\exp{-D}/\sqrt{D}$.)
The $\phi^{(k)}$ is a parameter that describes the background covariance within the layer. 
If two samples draw 100\% of their ancestry from layer $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the layer.
The ``shared drift'' parameter $\phi^{(k)}$ is analogous to 
the branch length connecting the $k$th population to the population ancestral to all modeled
layers (see for example \cite{patterson_ancient_2012,peter_fstats}),
although they cannot be directly compared because 
we are modeling the allelic, rather than genetic, covariance. 
In \ref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

@@

\paragraph{Likelihood and inference}
If the allele frequencies at each locus
were independent between loci and multivariate normally distributed across populations, 
their allelic covariance $\widehat{\Omega}$ 
would be Wishart distributed with degrees of freedom equal to $L$, 
the number of loci genotyped.
We use this as a convenient approximation to the true distribution described above,
and so define the likelihood of the allelic covariance to be
\begin{equation}
    P(\widehat{\Omega} \; | \; \Omega) 
    = 
    \mathcal{W} \left( L\widehat{\Omega} \; | \; \Omega,L\right) ,
\end{equation}
where $\mathcal{W}$ is the Wishart likelihood function.
Statistical nonindependence between loci (linkage disequilibrium)
will decrease the effective
number of degrees of freedom.  
One possible solution, which we have not yet found necessary to implement,
would be to estimate an \emph{effective} number of loci 
by introducing a parameter to modify the given degrees of freedom 
and thereby informally model linkage between loci (e.g., \cite{EEMS}).

We estimate the values of the parameters of the model using a Bayesian approach.
Acknowledging the dependence of the parametric covariance matrix $\Omega$ on its constituent parameters
$w,\alpha,\phi,\eta,\gamma$ and on the (observed) geographic distances $D$ 
with the notation $\Omega(w,\alpha,\phi,\eta,\gamma,D)$,
we denote the posterior probability density of the parameters as:
\begin{equation}
P\left( w,\alpha,\phi,\eta,\gamma \;|\; \widehat{\Omega} \right) 
    \propto
    P\left(\widehat{\Omega} \; | \; \Omega(w,\alpha,\phi,\eta,\gamma,D) \right)
    P(w)P(\alpha)P(\phi)P(\eta)P(\gamma) ,
\end{equation}
where $P(w)$, $P(\alpha)$, $P(\phi)$, $P(\eta)$, and $P(\gamma)$, are prior distributions.
All parameters are given (half-)Gaussian priors 
except for $\alpha_2$, which is uniform on $(0,2)$, 
and $w$, for which we use an independent Dirichlet of dimension $K$ for each individual
(see Table \ref{tab:param_prior_tab} for specifics).
Parameters are independent between layers.
We use Hamiltonian Monte Carlo as implemented in STAN \cite{stan, NUTS, stan_lib, rstan} 
to estimate the posterior distribution on the parameters.
We also provide an R package \cite{R} called \texttt{conStruct} 
(for ``continuous structure") that functions as a wrapper around this inference machinery.

@@

A nice feature of our approach is that the model described in Eq.\ \eqref{cross_layer_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is very similar to that of STRUCTURE \cite{STRUCTURE} 
and related models (e.g., \cite{ADMIXTURE}); 
the \gc{main} difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \cite{Engelhardt2012}.  
(We make this approximation for the substantial advantages in
computational speed.)
\gc{The second small difference is that in the original STRUCTURE model,
  allele frequencies at each locus are independently drawn for each
  cluster \cite{STRUCTURE}, while in conStruct's non-spatial model,
  it is more natural to envision each layer's allele frequency
  as being drifted away from a single, global allele frequency. 
  This makes our model more
  closely related to the ``$F$-model'' prior for allele frequencies \cite{falush2003}.
As these differences are relatively small,} we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework. 
We also validate our claim that the results of our nonspatial method 
are a close match to those of STRUCTURE-based approaches.

@@

Several approaches have been used as model choice criteria 
for the number of discrete clusters in population genetic data, including: 
information theoretic criteria (e.g., \cite{ADMIXTURE}), such as AIC or BIC;
comparisons of the likelihood of the data across different values of $K$, 
with various criteria on how to choose a single value (e.g., \cite{Evanno2005});
or comparisons of the marginal likelihood, 
generated either via various approximations (e.g., \cite{STRUCTURE})
or via a more rigorous and computationally intensive approach such as thermodynamic integration \cite{verity_nichols2016}
or inference using a Dirichlet process prior \cite{huelsenbeck2007inference}.
See \cite{verity_nichols2016} for a discussion of these approaches and comparison
between several methods.

We use cross-validation (similar in spirit to \cite{ADMIXTURE_xval}) to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

@@

Another way to describe the practical significance of additional layers
is to calculate each layer's relative contribution to total covariance, 
and to choose a value of $K$ where all layers have a contribution above some cutoff (e.g. 0.1\%).
The Dirichlet prior on admixture proportions
is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
encouraging the model to ``not use" unnecessary layers if they are
present \gc{in the model'}, 
so that they will have a low contribution to overall covariance.

To calculate layer contributions, we use the following alternative description of our covariance model:
the genomes of any pair of individuals \emph{agree} with some
background probability \gc{at a locus},
but this probability of agreement is increased on any segment of genome 
that both have inherited \gc{from} the same layer 
(the amount it increases depends on how far apart they are
\gc{geographically and the decay of IBD}.)
We use this characterization to quantify the relative contributions of each layer,
by computing the average contribution to increased probability of agreement
as described in \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

To test the method, we first generated data using the coalescent simulator \texttt{ms} \cite{Hudson2002}.
In each simulation,
we split a single ancestral population into $K$ subpopulations $\tau_{\text{s}}$ units of coalescent time in the past,
and at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonized a
separate $6 \times 6$ square lattice of demes.
Migration on each lattice was to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past, 
we collapsed those $K$ discrete layers into a single grid of demes,
choosing various amounts of admixture from these different layers
(see Fig \ref{sim_setup}). 
We collapsed the layers together using random admixture with spatially autocorrelated proportions 
(see \secref{sim_details} for details), 
so that both the frequencies in layers and the layer memberships are
spatially autocorrelated. 
We simulated datasets using $K=1$, 2, and 3 layers;
in each simulation we sampled 10,000 unlinked loci from each of 20 haploid individuals from every deme.
We then ran both spatial and nonspatial \texttt{conStruct} analyses on each simulated dataset with $K$ between 1 and 7,
and compared predictive performance of the models using cross-validation.
% To confirm that results from our nonspatial model were qualitatively similar to those of 
% other nonspatial model-based clustering algorithms, 
For comparison,
we also analyzed each simulated dataset using fastSTRUCTURE \cite{fastStructure}, using $K=2-4$.
% inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}--\ref{fastStr_simK3}).

% In Fig \ref{simK1_pies}, we show the results of the dataset simulated with $K=1$, 
% contrasting the results of the nonspatial model runs with 2--4 layers 
% with those of spatial model with 2--4 layers.
% 
% In Fig \ref{sim_xvals} we show the results of the cross-validation analysis for each of the three simulated datasets.
% The results of all analyses run on datasets simulated using $K=1$ through 3 are shown in 
% Figs \ref{simK1_nsp_pies}-\ref{simK3_adprop_fit}.
% 
% To confirm that results from our nonspatial model were qualitatively similar to those of 
% other nonspatial model-based clustering algorithms, 
% we also analyzed each simulated dataset using fastSTRUCTURE \cite{fastStructure}, 
% inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}--\ref{fastStr_simK3}).

@@

Trees in the genus \textit{Populus} (poplars, aspens, and cottonwoods) 
are distributed throughout the Northern Hemisphere;
species in the genus regularly co-occur and, 
where they do, they frequently hybridize \cite{eckenwalder1984, Cronk2005}.

\textit{Populus trichocarpa}, the black cottonwood,
and \textit{Populus balsamifera}, the balsam poplar,
have a broad zone of overlap in the Pacific Northwest,
where they are hypothesized to hybridize \cite{geraldes_etal_2014, suarezgonzalez_etal_2016}.
Both species are sampled over a large geographic region, 
and show spatial patterns of genetic and phenotypic variation \cite{slavov_etal_2012, mckown_etal_2013},
making the system well-suited for application of our method.
% to test an analysis designed to describe
% how genetic structure is partitioned both discretely across populations (or species)
% and continuously with geographic distance.
We organize the results of our analyses around the following questions: 
\begin{enumerate}

@@

Across all values of $K$ for which we ran \texttt{conStruct},
we see strong support for the spatial model over the nonspatial model (Fig \ref{bear_xvals}).
% indicating the isolation by distance seems to be a prominent feature of the data.
This pattern may resolve a discrepancy between our results and 
previous analyses that split Alaskan and British Columbian bears out 
into their own cluster with an inferred Beringian glacial refugium
\cite{Byun1997,Stone2000,Puckett2015}.
Our model, which explicitly incorporates a spatial decay of relatedness, 
allows somewhat genetically differentiated individuals 
that are sampled far from one another to belong to the same layer,
instead of splitting these individuals out into successive clusters 
(e.g., Fig \ref{bear_sp_pies}d vs \ref{bear_nsp_pies}d).
